---
title: "Lecture 2: Investigating data patterns" # potentially push to header
subtitle:  "Managing and Manipulating Data Using R"
author: 
date: 
classoption: dvipsnames  # for colors
fontsize: 8pt
urlcolor: blue
output:
  beamer_presentation:
    keep_tex: true
    toc: false
    slide_level: 3
    theme: default # AnnArbor # push to header?
    #colortheme: "dolphin" # push to header?
    #fonttheme: "structurebold"
    highlight: tango # Supported styles include "default", "tango", "pygments", "kate", "monochrome", "espresso", "zenburn", and "haddock" (specify null to prevent syntax highlighting); push to header
    df_print: tibble #default # tibble # push to header?    
    latex_engine: xelatex #  Available engines are pdflatex [default], xelatex, and lualatex; The main reasons you may want to use xelatex or lualatex are: (1) They support Unicode better; (2) It is easier to make use of system fonts.
    includes:
      in_header: ../beamer_header.tex
      #after_body: table-of-contents.txt 
---

```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
#knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```

```{r, echo=FALSE, include=FALSE}
#DO NOT WORRY ABOUT THIS
if(!file.exists('transform-logical.png')){
  download.file(url = 'https://raw.githubusercontent.com/ozanj/rclass/master/lectures/lecture2/transform-logical.png', 
                destfile = 'transform-logical.png',
                mode = 'wb')
}
```


```{r echo=FALSE}
#DO NOT WORRY ABOUT THIS
if(!file.exists("fp1.JPG")) {
  download.file(url ="https://github.com/ozanj/rclass/raw/master/lectures/lecture2/fp1.JPG",
                destfile = 'fp1.JPG',
                mode = 'wb')
}
```

```{r echo=FALSE}
#DO NOT WORRY ABOUT THIS
if(!file.exists("fp2.JPG")) {
  download.file(url ="https://github.com/ozanj/rclass/raw/master/lectures/lecture2/fp2.JPG",
                destfile = 'fp2.JPG',
                mode = 'wb')
}
```

# Introduction

### What we will do today

\tableofcontents
```{r, eval=FALSE, echo=FALSE}
#Use this if you want TOC to show level 2 headings
\tableofcontents
#Use this if you don't want TOC to show level 2 headings
\tableofcontents[subsectionstyle=hide/hide/hide]
```

### Logistics

1. Volunteer teaching assistant, Diana Lopez
1. New student introductions
1. Homework expectations
    - due date: before class (except first homework due today)
    - general instructions for homework
    - grading (enrolled students, auditing students)
1. Added required reading for next week on syllaus 
1. If we don't get through all of lecture, finish on your own
    - whenever we don't cover difficult concepts in class; I'll cover those concepts next week and slow down the pace of class
    
# Find homework groups

### What we'll do to choose homework groups

- Meet new people (10 minutes of speed-dating!)
    - find someone in class you don't know and talk to them for two minutes about anything
        - e.g., where you from, what program, what are research interests, what you like doing outside of school, work
- Enrolled students choose homework groups (10 minutes)
    - one side of room for students who want to work collaboratively on problem sets
    - one side of room for students who want to work mostly on their own (e.g,. due to full work/family schedule);
    - must be groups of 3
    - cannot have more than 2 people from same academic program (e.g., HEOC, HDP)
- Auditors not part of "official" homework groups of 3, but they are welcome to join any homework group or form their own homework group

Recommendation: **Use Zoom for group meetings!**

- https://ucla.zoom.us/ 