---
title: "Problem Set #1 Solutions"
author: "Patricia Martin"
date: "9/28/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem set 1  

### Step 1:

- Go to your rclass folder and open your saved lecture1.2.Rmd file (hint: in lecture1 folder)  
- Open the lecture1.2.Rmd file in RStudio  
- At the top of the lecture1.2.Rmd file, before the "# R basics", type your first and last name (e.g. "Patricia Martin")  
- Now select the "Knit" tab (icon with blue yarn ball) or scroll down and "Knit to PDF (Beamer)"  
- The goal of this assignment is to make sure you are able to knit to a Beamer presentation (PDF) without running into errors.  


### Step 2:  

**Question 1**: Load .Rdata directly with url  

- Make sure to view current/working directory first `getwd()` 
- Url to data https://github.com/ozanj/rclass/raw/master/data/recruiting/recruit_ps1_small.Rdata  
- Hint: to load .Rdata use `load()` function and "url" because you are using a link  
- E.g. `load(url(link))`  

```{r}
getwd()
?load
load(url("https://github.com/ozanj/rclass/raw/master/data/recruiting/recruit_ps1_small.Rdata"))
```


**Question 2**: Set your working directory and use absolute filepath to load data  

- Download .Rdata using link from above  
- Save .Rdata file in the recruiting folder (rclass/data/recruiting)  
- View your current/working directory `getwd()` 
- Use `setwd()` function to set your working directory to where .Rdata file is saved (rclass/data/recruiting)  
- When setting working directory, use the absolute filepath  
- Load the .Rdata file  `load(name.Rdata)`
- Using the `names(data frame name)` function, report the names in this dataset  (Hint: data frame name is df_recruiting)  

```{r}
getwd() 
setwd("/Users/lizn/rclass/data/recruiting") #absolute file path to data
getwd()  #if you run lines 50-52 together you can see how the working directory changes
list.files() #list.files displays the files in the folder
load("recruit_ps1_small.Rdata")
names(df_recruiting)
```

**Question 3**: Set your working directory and use relative filepath to load data  

- View your current/working directory `getwd()` 
- Use `setwd()` function to set your working directory to where .Rdata file is saved (rclass/data/recruiting)  
- When setting working directory, use the relative filepath  
- Load the .Rdata file  `load()`
- Report the names in this dataset (Hint: data frome name is df_recruiting)  

```{r}
getwd() 
# Working directory is in the lecture1 folder, need to get to recruiting folder using relative file path
setwd("../../data/recruiting")
getwd() #View current directory after setting the directory, run lines 67-70 
list.files() #listing files in the recruiting folder
# Lets load recruit_ps1_small.Rdata 
load("recruit_ps1_small.Rdata")
# Use names function to view the names in the data set (data frame)
names(df_recruiting)
# Run the entire code chunck 
```

**Question 4**: Load dataset using relative filepath without first changing working directory   


```{r}
getwd()
load("../../data/recruiting/recruit_ps1_small.Rdata") #Set file path and load data at once
names(df_recruiting)
```

**Knit to pdf** select the "Knit" tab (icon with blue yarn ball) or scroll down and "Knit to PDF"  

- Submit both .Rmd and pdf to rclass website https://ozanj.github.io/rclass/resources/ under week 1  
- Use this naming convention "lastname_firstname_ps1"