---
title: "homework 1"
author: Ozan Jaquette
urlcolor: blue
output: 
  html_document:
    toc: true

---

Should probably create a separate document of instructions for homework in general. we'll update this over time.

Second part of homework tells them to load some data in different ways. in the instructions for second part:

- provide them links to: (A) the basic youtube video on relative and absolute directories and (B) the youtube video about changing directories in R
    - tell them to re-watch these videos
- tell them about list.files() command
- tell them that `getwd()` is command to tell them filepath of current working directory
    - tel them that when you `getwd()` from a code chunk in .Rmd, the file-path will be the directory where the .Rmd file is stored
- tell them what `setwd()` does
- tell them that `setwd()` executed within a code chunk only changes working directory for that code chunk, then reverts back to previous working directory




some question where they load dataset directly from url we provide them [doesn't seem to be up yet]
```{r}
getwd()
```

Some question where they first change directories using `setwd()` using absolute filepath
```{r}
getwd() # in instructions, make sure that 
setwd("C:/Users/ozanj/Documents/rclass/data/recruiting")
getwd()
list.files()
load("recruit_event_somevars.Rdata")
names(df_event)
```

```{r}
load("C:/Users/ozanj/Documents/rclass/data/recruiting/recruit_event_somevars.Rdata")
rm(df_event)
```

Some question where they first change directories using `setwd()` using relative filepath
```{r}
getwd() # in instructions, make sure that 
setwd("../../data/recruiting")
getwd()
list.files()
load("recruit_event_somevars.Rdata")
names(df_event)
```

Some question where they load dataset using relative file-path without first changing directories
```{r}
getwd()
load("../../data/recruiting/recruit_event_somevars.Rdata")
names(df_event)
```

```{r}
getwd()
load("../../data/recruiting/recruit_ps1_small.Rdata")
names(df_recruiting)
```

