---
title: "Problem Set #2 Solutions"
author: "INSERT YOUR NAME HERE"
date: "10/4/2019"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
---

```{r, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```
<span style="color:red">Grade: /20</span>  

In this problem set, you will investigate objects and data patterns. 

# Step 1: Make changes to YAML header  
<span style="color:red">(/1)</span>

Recommend reading R Markdown: The Definative Guide section 3.1 ([LINK HERE](https://bookdown.org/yihui/rmarkdown/html-document.html)) before answering these questions

1. Add a table of contents to YAML header
1. table of contents should have "depth" of 2
1. table of contents should "float"
1. add section numbering to headers


Load tidyverse package  
<span style="color:red">(/0.5)</span>
```{r}
#install.packages("tidyverse") #install if you do not have tidyverse installed
library(tidyverse)
```

Load data frame objects
```{r}
rm(list = ls()) # remove all objects
getwd()

#glimpse(df_school)
load(url("https://github.com/ozanj/rclass/raw/master/data/recruiting/recruit_school_allvars.RData"))
#glimpse(df_school_all)

df_school_all %>% count(total_visits)
```


# Step 2: Investigating objects, Base R approach

1. Answer the following questions about the object `df_school_all` by running the appropriate R command within text (as opposed to in an R code chunk) and writing any substantive response required by the question. The first question will be answered for you to show how it works  
<span style="color:red">(/2)</span>  
    + What "type" of object is `df_school_all`?
        + __ANSWER [GIVEN]: the object `df_school_all` has type equals `r typeof(df_school_all)`__
    + What is the "length" of the object `df_school_all`? What does this specific value of length refer to?
        + __ANSWER: the length of object `df_school_all` is `r length(df_school_all)`. This value refers to the number of variables__
    + how many "rows" are in the object `df_school_all`? what does each row represent?
        + __ANSWER: the object `df_school_all` has `r nrow(df_school_all)` rows. Each row represents a high school__

```{r}
#type of df_school_all 
typeof(df_school_all)

#length of df_school_all 
length(df_school_all)

#num of rows in df_school_all
nrow(df_school_all)
```



\newline
2. In the below code chunk, use the `str()` function to describe the contents of `df_school_all` and then answer the following questions in text above the code chunk:  
<span style="color:red">(/2)</span>
    + What does each element of the object `df_school_all` represent? (hint Lecture 1.2, Slide 34)
        + __ANSWER: a variable__
    + Are the individual elements within `df_school_all` lists or vectors?
        + __ANSWER: vectors__    
    + Are the individual elements within `df_school_all` named or un-named? what do these element names refer to?
        + __ANSWER: the individual elements are named; the name of each element refers to a variable name__    
        
```{r}
str(df_school_all)
```

\newline

3. These questions refer to the variable `school_type` within the object `df_school_all`. For the first two questions, run the appropriate R command within text (as opposed to in an R code chunk) and write any substantive response required by the question  
<span style="color:red">(/1.5)</span>
    + What is the data "type" of `school_type`?
        + __ANSWER: The variable `school_type` has data type equal to `r typeof(df_school_all$school_type)`__
    + What is the "length" of `school_type`? What does this specific value of length refer to?
        + __ANSWER: The length of `school_type` is `r length(df_school_all$school_type)`. This value refers to the number of rows (or observations)__
        
```{r}
typeof(df_school_all$school_type)

length(df_school_all$school_type)
```
        
        
4. In these questions, you will apply the `table()` function to the variable `school_type` within the object `df_school_all`.  
<span style="color:red">(/3)</span>
    + In your own words, what does the `table()` function do?  
        + __ANSWER: counts the number of observations for each value of a variable__
    + What does the `useNA` argument of the `table()` function control?
        + __ANSWER: determines whether to include `NA` values in the table of counts__
    + What is the default value of the `useNA` argument and what does this default value mean?
        + __ANSWER: the default value for `useNA` argument is "no", which means that `NA` values are excluded from the count__
    + What happens when you assign the value `"ifany"` to the `useNA` argument?
        + __ANSWER: if there are observations with `NA` values, the table of counts shows the number of observations with `NAs`. If no observations with `NA` values, the table does not have a category for `NAs`__
    + What happens when you assign the value `"always"` to the `useNA` argument?
        + __ANSWER: The table of counts shows the number of `NA` observations even if there are zero `NA` observations__
    + In the below R code chunk, use the `table()` function to count the number of observations for each value `school_type` three different ways: 
        + First, without specifying any value for `useNA`
        + Second, by assigning the value `"ifany"` to the `useNA` argument
        + Third,  by assigning the value `"always"` to the `useNA` argument
```{r}
table(df_school_all$school_type)
table(df_school_all$school_type, useNA = "ifany")
table(df_school_all$school_type, useNA = "always")
```

# Filter, select, arrange questions

The data frame `df_school_all` has one observation for each high school (public and private). 

- The variables that begin with `visits_by_...` identify how many off-campus recruiting visits the high school received from a particular public university. For example UC Berkeley has the ID `110635` so the variable `visits_by_110635` identifies how many visits the high school received from UC Berkeley. 
- The variable `total_visits` identifies the number of visits the high school received from all (16) public research universities in this data collection sample.

For the questions below, imagine that you have been asked by a major news outlet to identify which high schools receive the most total number of off-campus recruiting visits from public universities. 

- for all questions below, you can answer using one line of code or you can answer in several steps (e.g., first create new data frame, then print selected variables)
- For questions that ask you to print the "top 10" observations, you can simply print the object and rely on the fact that the default option [for "tibble" data frames] is to print 10 observations OR you can wrap the command in the `head()` function and explicitly tell R to print 10 observations
- Before conducting analyses, we'll rename the variable `avgmedian_inc_2564` to give it a shorter name.

1. Rename the variable `avgmedian_inc_2564` to `med_inc` and assign new variable name to the existing object `df_school_all`  
<span style="color:red">(/2)</span>
```{r}
df_school_all <- rename(df_school_all, med_inc = avgmedian_inc_2564)
#names(df_school_all)
```


2. The news outlet is interested in comparing the in-state and out-of-state high school visits for [The University of Alabama](https://nces.ed.gov/globallocator/col_info_popup.asp?ID=100751) (IPEDS ID = 100751) variable `visits_by_100751`. Count the number of in-state public high schools that received at least one visit from The University of Alabama.  
<span style="color:red">(/2)</span>
    - Note: You will need to use `filter` and the `count` function  
    - Use commas to separate varibles for this questions (e.g. `filer(dataframe, variable == something, variable == something)`)
    - You can do this in one step by wrapping the `count` function around the `filter` function ; or you can do this in two steps by creating a new data frame first  
```{r}
count(filter(df_school_all, visits_by_100751 >= 1, state_code == "AL", school_type== "public"))

```

3. How many public in-state high schools visited by The University of Alabama enroll at least 50% Latinx **or** 50% Black students?  
<span style="color:red">(/2)</span>
```{r}
count(filter(df_school_all, school_type == "public", pct_black >= 50 | pct_hispanic >= 50,
             state_code == "AL", visits_by_100751 >= 1))
```


4. Now count the out-of-state public high schools that received at least one visit by the University of Alabama without using commas.  
<span style="color:red">(/2)</span>
    - hint: `&` or `%in%`
```{r}
count(filter(df_school_all, visits_by_100751 >=1 & state_code != "AL" & school_type == "public"))

```

5. How many public out-of-state high schools visited by The University of Alabama enroll at least 50% Latinx **or** 50% Black students?  
<span style="color:red">(/2)</span>
```{r}
count(filter(df_school_all, visits_by_100751 >= 1, school_type == "public", state_code != "AL" , pct_black >= 50 | pct_hispanic >= 50))
```


Once finished, knit to (HTML) and upload both .Rmd and HTML files to ps_submit_2 on Piazza

*Remember to use this naming convention "lastname_firstname_ps2"*

