---
title: "Problem Set Resources and Guideline"
date: "9/27/2019"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Resources 
## **1.** Re-watch these videos on working with relative and absolute paths  

  * Absolute versus relative file paths [Youtube link](https://www.youtube.com/watch?v=ephId3mYu9o)  
  * Relative paths and working directory in R [Youtube link](https://www.youtube.com/watch?v=fe6GA200dks)  
  
## **2.** `list.files()` command  
  * The list.files command will list the files in the current working directory  
  * To learn more type `?list.files` in the console 
```{r}
?list.files
list.files() #currently in rclass project folder
```
  * Alternatively, you could set the path to the folder you would like to list files
```{r}
list.files(path = "data/ipeds")
```
## **3.** To view the filepath of your current/working directory use the `getwd()` function.  
  * Note that when you use the `getwd()` function from a code chunk in .Rmd, the filepath will be the directory where the .Rmd file is saved. 
```{r}
getwd()
```
  * This pdf "problemset_instructions.Rmd" is saved in the rclass folder (current working directory)  
  
## **4.** To set your working directory, use the `setwd()` function.  
  * Note that when `setwd()` is executed within a code chunk, the working directory changes only for that code chunk and will revert back to previous working directory in another code chunk.  
  
```{r}
getwd()
setwd("lecture2")
```

## **5.** Once problem sets are complete, knit to pdf or html (depending on instructions)  

- Select the "Knit" tab (icon with blue yarn ball) or scroll down and "Knit to PDF" or "Knit to HTML" 
- Go to class website https://ozanj.github.io/rclass/resources/ and each week collapsible will have a link titled "Sumbit Problem Set Here"  
- Select the link and submit both .Rmd and pdf/html files  
- Make sure to use this naming convention "lastname_firstname_ps#" 

## **6.** Problem sets  

- All problem sets are worth the same. The lowest problem set grade will be dropped from the calculate of your final grade. 

- Submit problem sets on Piazza. Link to Piazza: https://piazza.com/ucla/fall2019/educ263/home

    - Select "Q&A" at the top
    
    - Select "New Post" tab on the left corner
    
    - Post Type: Note
    
    - Post to: select "Instructor(s)", delete the "Instructors" tag and enter "Xin Li" (Be sure NOT post to the entire class but just TA)
    
    - Select Folder(s): ps_submit_1
    
    - Summary: using this naming convention "lastname_firstname_ps1"
    
    - Details: insert lastname_firstname_ps1.Rmd & lastname_firstname_ps1.pdf file
    
    - At the bottom, you can preview your post before posting
    
    - Once you are ready, you can post your note
    
    - TA will give feedback to your post
    
    
## **7.** Homework Groups  
- \textcolor{red}{\textbf{Clarification:}} Each student is responsible for turning in their own problem set (as opposed to turning in one problem set per group). We, however, strongly encourage you all to work in your homework groups on the problem sets. We recommend that you attempt to work through the problem sets prior to meeting with your group, but that is only a recommendation, not a requirement.


