---
title: "Problem Set Resources and Guideline"
date: "9/24/21"
output:
  pdf_document: default
  word_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Resources 
## **1.** Re-watch these videos on working with relative and absolute paths  

- Absolute versus relative file paths [Youtube link](https://www.youtube.com/watch?v=ephId3mYu9o)  
- Relative paths and working directory in R [Youtube link](https://www.youtube.com/watch?v=fe6GA200dks)  
  
## **2.** `list.files()` command  

- The list.files command will list the files in the current working directory  
- To learn more type `?list.files` in the console  

```{r}
?list.files
list.files() #currently in rclass project folder
```

- Alternatively, you could set the path to the folder you would like to list files
```{r}
list.files(path = "data/ipeds")
```
## **3.** To view the filepath of your working directory use the `getwd()` function.  

- Note that when you use the `getwd()` function from a code chunk in .Rmd, the filepath will be the directory where the .Rmd file is saved. 
```{r}
getwd()
```

- This pdf "problemset_instructions.Rmd" is saved in the problemset_resources folder (current working directory)  
  
## **4.** To set your working directory, use the `setwd()` function.  

- Note that when `setwd()` is executed within a code chunk, the working directory changes only for that code chunk and will revert back to previous working directory in another code chunk.  
  
```{r}
getwd()
setwd("../intro_to_r")
```

## **5.** Once problem sets are complete, knit to pdf 

- Select the "Knit" tab (icon with blue yarn ball) or scroll down and "Knit to PDF" or "Knit to HTML"  
- Go to class website https://anyone-can-cook.github.io/rclass1/ in the "Readings & Assignments" section and under each week, there will be a link titled "Problem set submission link"  
- Click on the link and submit both .Rmd and pdf/html files  
- Make sure to use this naming convention "lastname_firstname_ps#"


## **6.** Questions & Class discussion  

- We encourage you to post any questions related to the problem sets and/or lectures as a GitHub issue. You can post issues here https://github.com/anyone-can-cook/rclass1_student_issues_f21.  

- **Issue guidelines**  

  - Clear title describing your issue (e.g. PS1: file directory question)  
  - Provide enough background information about your issue including a description, your code, any relevant links, etc.  
  - Assign and/or mention `@` the instructors (@ozanj, @lizachavac, @briannawright135) in your issue.  
  - Assign and/or mention members of your group, if applicable.

    
    
## **7.** Homework Groups  
- \textcolor{red}{\textbf{Clarification:}} Each student is responsible for turning in their own problem set (as opposed to turning in one problem set per group). We, however, strongly encourage you all to work in your homework groups on the problem sets. We recommend that you attempt to work through the problem sets prior to meeting with your group, but that is only a recommendation, not a requirement.


