---
title: "Lecture 7: Guidelines for Investigating, cleaning, and creating variables" # potentially push to header
subtitle:  "Managing and Manipulating Data Using R"
author: 
date: 
classoption: dvipsnames  # for colors
fontsize: 8pt
urlcolor: blue
output:
  beamer_presentation:
    keep_tex: true
    toc: false
    slide_level: 3
    theme: default # AnnArbor # push to header?
    #colortheme: "dolphin" # push to header?
    #fonttheme: "structurebold"
    highlight: tango # Supported styles include "default", "tango", "pygments", "kate", "monochrome", "espresso", "zenburn", and "haddock" (specify null to prevent syntax highlighting); push to header
    df_print: tibble #default # tibble # push to header?    
    latex_engine: xelatex #  Available engines are pdflatex [default], xelatex, and lualatex; The main reasons you may want to use xelatex or lualatex are: (1) They support Unicode better; (2) It is easier to make use of system fonts.
    includes:
      in_header: ../beamer_header.tex
      #after_body: table-of-contents.txt 
---

```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
#knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```


# Introduction

### What we will do today

\tableofcontents


### Libraries 

"Load" the package we will use today (output omitted)

- __you must run this code chunk after installing these packages__
```{r, message=FALSE}
library(tidyverse)
library(haven)
library(labelled)
```
__If package not yet installed__, then must install before you load. Install in "console" rather than .Rmd file

- Generic syntax: `install.packages("package_name")`
- Install "tidyverse": `install.packages("tidyverse")`

Note: when we load package, name of package is not in quotes; but when we install package, name of package is in quotes:

- `install.packages("tidyverse")`
- `library(tidyverse)`


### Data 

Use `read_dta()` function from `haven` to import Stata dataset into R
```{r, results="hide"}
hsls <- read_dta(file="https://github.com/ozanj/rclass/raw/master/data/hsls/hsls_stu_small.dta")
```

Let's examine the data [you __must__ run this code chunk]
```{r, results="hide"}
hsls %>% names()
hsls %>% names() %>% str()
hsls %>% names() %>% tolower() %>% str()

names(hsls) <- tolower(names(hsls)) # convert names to lowercase
names(hsls)

str(hsls) # ugh

str(hsls$s3classes)
attributes(hsls$s3classes)
typeof(hsls$s3classes)
class(hsls$s3classes)
```

Download the HSLS Codebook: https://nces.ed.gov/pubs2014/2014361_AppendixI.pdf

# Exploratory data analysis (EDA)

### What is exploratory data analysis (EDA)?

The [Towards Data Science](https://towardsdatascience.com/exploratory-data-analysis-8fc1cb20fd15) website has a nice definition of EDA:

> "Exploratory Data Analysis refers to the critical process of performing initial investigations on data so as to discover patterns,to spot anomalies,to test hypothesis and to check assumptions with the help of summary statistics" 

__This course focuses on "data management":__

- investigating and cleaning data for the purpose of creating analysis variables
- Basically, everything that happens __before__ you conduct analyses

__I think about "exploratory data analysis for data quality"__

- Investigating values and patterns of variables from "input data"
- Identifying and cleaning errors or values that need to be changed
- Creating analysis variables
- Checking values of analysis variables agains values of input variables

### How we will teach exploratory data analysis

Will teach exploratory data analysis (EDA) in two sub-sections:

1. Introduce "Tools of EDA": 
    - Demonstrate code to investigate variables and relatioship between variables
    - Most of these tools are just the application of programming skills you have already learned
1. Provide "Guidelines for EDA"
    - Less about coding, more about practices you should follow and mentality necessary to ensure high data quality

## Tools for EDA


### Tools of EDA

__To do EDA for data quality, must master the following tools:__

- \medskip __Select, sort, filter, and print__ in order to see data patterns, anomolies
    - Select and sort particular values of particular variables
    - Print particular values of particular variables
- __One-way descriptive analyses__ (i.e,. focus on one variable)
    - Descriptive analyses for continuous variables
    - Descriptive analyses for discreet/categorical variables
- __Two-way descriptive analyses__ (relationship between two variables)
    - Categorical by categorical
    - Categorical by continuous
    - Continuous by continuous

Whenever using any of these tools, __pay close attention to missing values and how they are coded__

- Often, the "input" variables don't code missing values as `NA` 
- Especially when working with survey data, missing values coded as a negative number (e.g., `-9`,`-8`,`-4`) with different negative values representing different reasons for data being missing
- sometimes missing values coded as very high positive numbers
- Therefore, important to investigate input vars prior to creating analysis vars

### Tools of EDA

First, Let's create a smaller version of the HSLS:09 dataset
```{r}
#hsls %>% var_label()
hsls_small <- hsls %>%
  select(stu_id,x3univ1,x3sqstat,x4univ1,x4sqstat,s3classes,
         s3work,s3focus,s3clgft,s3workft,s3clgid,s3clgcntrl,
         s3clglvl,s3clgsel,s3clgstate,s3proglevel,x4evrappclg,
         x4evratndclg,x4atndclg16fb,x4ps1sector,x4ps1level,
         x4ps1ctrl,x4ps1select,x4refsector,x4reflevel,x4refctrl,
         x4refselect, x2sex,x2race,x2paredu,x2txmtscor,x4x2ses,x4x2sesq5)
```

```{r, results="hide"}
names(hsls_small)
hsls_small %>% var_label()
```


### Tools of EDA: select, sort, filter, and print

We've already know `select()`, `arrange()`, `filter()` 

\medskip Select, sort, and print specific vars
```{r, results="hide"}
#sort and print
hsls_small %>% arrange(desc(stu_id)) %>% 
  select(stu_id,x3univ1,x3sqstat,s3classes,s3clglvl)

#investigate variable attributes
hsls_small %>% arrange(desc(stu_id)) %>% 
  select(stu_id,x3univ1,x3sqstat,s3classes,s3clglvl) %>% str()

#print observations with value labels rather than variable values
hsls_small %>% arrange(desc(stu_id)) %>% 
  select(stu_id,x3univ1,x3sqstat,s3classes,s3clglvl) %>% as_factor()
```
Sometimes helpful to increase the number of observations printed
```{r, results="hide"}
class(hsls_small) #it's a tibble, which is the "tidyverse" version of a data frame
options(tibble.print_min=50) 
# execute this in console
hsls_small %>% arrange(desc(stu_id)) %>%
  select(stu_id,x3univ1,x3sqstat,s3classes,s3clglvl)
options(tibble.print_min=10) # set default printing back to 10 lines
```

### One-way descriptive stats for continuous vars, Base R approach [SKIP]

```{r, results="hide"}
mean(hsls_small$x2txmtscor)
sd(hsls_small$x2txmtscor)

#Careful: summary stats include value of -8!
min(hsls_small$x2txmtscor)
max(hsls_small$x2txmtscor)
```
Be careful with `NA` values
```{r, results="hide"}
#Create variable replacing -8 with NA
hsls_small_temp <- hsls_small %>% 
  mutate(x2txmtscorv2=ifelse(x2txmtscor==-8,NA,x2txmtscor))
hsls_small_temp %>% filter(is.na(x2txmtscorv2)) %>% count(x2txmtscorv2)

mean(hsls_small_temp$x2txmtscorv2)
mean(hsls_small_temp$x2txmtscorv2, na.rm=TRUE)
rm(hsls_small_temp)
```
### One-way descriptive stats for continuous vars, Tidyverse approach

Use `summarise_at()`, a variation of `summarise()`, to make descriptive stats

- `.args=list(na.rm=TRUE)`= a named list of additional arguments to be added to all function calls

__Task__:

- calculate descriptive stats for `x2txmtscor`, math test score
```{r, warning=FALSE}
#?summarise_at
hsls_small %>% select(x2txmtscor) %>% var_label()
hsls_small %>% 
  summarise_at(
    .vars = vars(x2txmtscor),
    .funs = funs(mean, sd, min, max, .args=list(na.rm=TRUE))
  )
```

### One-way descriptive stats for continuous vars, Tidyverse approach

Can calculate descriptive stats for more than one variable at a time

__Task__:

- calculate descriptive stats for `x2txmtscor`, math test score, and `x4x2ses`, socioeconomic index score


```{r}
hsls_small %>% select(x2txmtscor,x4x2ses) %>% var_label()

hsls_small %>% 
  summarise_at(
    .vars = vars(x2txmtscor,x4x2ses),
    .funs = funs(mean, sd, min, max, .args=list(na.rm=TRUE))
  )
```

### One-way descriptive stats for continuous vars, Tidyverse approach

"Input vars" in survey data often have negative values for missing/skips
```{r, results="hide"}
hsls_small %>% filter(x2txmtscor<0) %>% count(x2txmtscor)
```

R includes those negative values when calculating stats; you don't want this

- Solution: create version of variable that replaces negative values with `NA`
```{r}
hsls_small %>% mutate(x2txmtscor_na=ifelse(x2txmtscor<0,NA,x2txmtscor)) %>%
  summarise_at(
    .vars = vars(x2txmtscor_na),
    .funs = funs(mean, sd, min, max, .args=list(na.rm=TRUE))
  )
```
What if you didn't include `.args=list(na.rm=TRUE)`?
```{r}
hsls_small %>% mutate(x2txmtscor_na=ifelse(x2txmtscor<0,NA,x2txmtscor)) %>%
  summarise_at(
    .vars = vars(x2txmtscor_na),
    .funs = funs(mean, sd, min, max))
```


### One-way descriptive stats for continuous vars, Tidyverse approach

How to identify these missing/skip values if you don't have a codebook?

- `count()` combined with `filter()` helpful for finding extreme values of continuous vars, which are often associated with missing or skip

```{r}
#variable x2txmtscor
hsls_small %>% filter(x2txmtscor<0) %>% 
  count(x2txmtscor)

#variable s3clglvl
hsls_small %>% select(s3clglvl) %>% var_label()

hsls_small %>% filter(s3clglvl<0) %>%
  count(s3clglvl)
```

### One-way descriptive stats student exercise

1. Using the object `hsls`, identify variable type, variable class, and check the variable values and value labels of `x4ps1start`
    - variable `x4ps1start` identifies month and year student first started postsecondary education
    - **Note**: This variable is a bit counterintuitive.
        - e.g., the value `201105` refers to May 2011
2. Get a frequency count of the variable `x4ps1start`  
3. Get a frequency count of the variable, but this time only observations that have negative values __hint__: use filter()  
4. Create a new version of the variable `x4ps1start_na` that replaces negative values with NAs and use `summarise_at()` to get the min and max value.  



### One-way descriptive stats student exercise solutions
\medskip

1. Using the object `hsls`, identify variable type, variable class, and check the variable vakyes and value labels of `x4ps1start`
```{r}
typeof(hsls$x4ps1start)
class(hsls$x4ps1start)

hsls %>% select(x4ps1start) %>% var_label()

hsls %>% select(x4ps1start) %>% val_labels()
```

### One-way descriptive stats student exercise solutions  
2. Get a frequency count of the variable `x4ps1start` 
```{r}
hsls %>%
  count(x4ps1start)
```

### One-way descriptive stats student exercise solutions
3. Get a frequency count of the variable, but this time only observations that have negative values __hint__: use filter()  
```{r}
hsls %>% 
  filter(x4ps1start<0) %>% 
  count(x4ps1start)
```

### One-way descriptive stats student exercise solutions
4. Create a new version `x4ps1start_na` of the variable `x4ps1start` that replaces negative values with NAs and use `summarise_at()` to get the min and max value.
```{r}
hsls %>% mutate(x4ps1start_na=ifelse(x4ps1start<0,NA,x4ps1start)) %>%
  summarise_at(
    .vars = vars(x4ps1start_na),
    .funs = funs(min, max, .args=list(na.rm=TRUE))
  )
```

### One-way descriptive stats for discrete/categorical vars, Tidyverse approach

Use `count()` to investigate values of discrete or categorical variables

For variables where `class==labelled`
```{r, results="hide"}
class(hsls_small$s3classes)
attributes(hsls_small$s3classes)
#show counts of variable values
hsls_small %>% count(s3classes) #print in console to show both
#show counts of value labels
hsls_small %>% count(s3classes) %>% as_factor()
```

- I like `count()` because the default setting is to show `NA` values too!
```{r, results="hide"}
hsls_small %>% mutate(s3classes_na=ifelse(s3classes<0,NA,s3classes)) %>% 
  count(s3classes_na)
```

Simultaneously show both values and value labels on count tables for `class==labelled` if entered into console

- This requires some concepts/functions we haven't introduced [SKIP]
```{r, eval=FALSE}
 hsls_small %>% count(s3classes)
y <- hsls_small %>% count(s3classes) %>% as_factor()
 bind_cols(x[,1], y) #wont show in updated R
```

<!-- ### One-way descriptive stats for factor vars [OPTIONAL/SKIP] -->

<!-- For variables where `class==factor` -->

<!-- - Create object with some or all `labelled` vars converted to `factor` -->
<!-- ```{r} -->
<!-- hsls_f <- as_factor(hsls) -->
<!-- ``` -->



<!-- ```{r, results="hide"} -->
<!-- #use variable from the hsls data frame where vars are factors -->
<!-- typeof(hsls_f$s3classes) -->
<!-- class(hsls_f$s3classes) -->
<!-- attributes(hsls_f$s3classes) -->

<!-- #show frequency table -->
<!-- hsls_f %>% count(s3classes) -->

<!-- #Create VAR that converts different types of missing to NA and then create frequency table -->

<!-- #note: within ifelse() used levels(s3classes)[s3classes]) rather than s3classes  to show attribute levels not values -->
<!-- hsls_f %>%  -->
<!--   mutate(s3classes_f=ifelse(s3classes %in% c("Missing","Unit      -->
<!--   non-response"),NA,levels(s3classes)[s3classes])) %>%  -->
<!--   count(s3classes_f) -->
<!-- ``` -->


### Relationship between variables, categorical by categorical

Two-way frequency table, called "cross tabulation", important for data quality

- When you create categorical analysis var from single categorical "input" var  
    - Two-way tables show us whether we did this correctly  
- Two-way tables helpful for understanding skip patterns in surveys

__key to syntax__

- `df_name %>% group_by(var1) %>% count(var2)` **OR**
- `df_name %>% count(var1,var2)`
- play around with which variable is `var1` and which variable is `var2`


### Relationship between variables, categorical by categorical


__Task__: Create a two-way table between `s3classes` and `s3clglvl`

- Investigate variables
```{r, results="hide"}
hsls_small %>% select(s3classes,s3clglvl) %>% var_label()
hsls_small %>% select(s3classes,s3clglvl) %>% val_labels()
```

- Create two-way table
```{r, results = "hide"}
hsls_small %>% group_by(s3classes) %>% count(s3clglvl) # show values
hsls_small %>% count(s3classes,s3clglvl)
#hsls_small %>% group_by(s3classes) %>% count(s3clglvl) %>% as_factor() # show value labels
```
- Are these objects the same?
```{r}
hsls_small %>% group_by(s3classes) %>% count(s3clglvl) %>% glimpse()
hsls_small %>% count(s3classes,s3clglvl) %>% glimpse()
```

### Relationship between variables, categorical by categorical

Two-way frequency table, also called "cross tabulation"

<!-- What if one of the variables has `NAs`? -->

<!-- - Table created by `group_by()` and `count()` shows `NAs`! -->

__Task__: 

- Create a version of `s3classes` called `s3classes_na` that changes negative values to `NA`
- Create a two-way table between `s3classes_na` and `s3clglvl`
```{r, results="hide"}
hsls_small %>% 
  mutate(s3classes_na=ifelse(s3classes<0,NA,s3classes)) %>%
  group_by(s3classes_na) %>% count(s3clglvl)

hsls_small %>% 
  mutate(s3classes_na=ifelse(s3classes<0,NA,s3classes)) %>%
  count(s3classes_na, s3clglvl)


#example where we create some NA obs in the second variable
hsls_small %>% 
  mutate(s3classes_na=ifelse(s3classes<0,NA,s3classes),
         s3clglvl_na=ifelse(s3clglvl==-7,NA,s3clglvl)) %>%
  group_by(s3classes_na) %>% count(s3clglvl_na)


hsls_small %>% 
  mutate(s3classes_na=ifelse(s3classes<0,NA,s3classes),
         s3clglvl_na=ifelse(s3clglvl==-7,NA,s3clglvl)) %>%
  count(s3classes_na, s3clglvl_na)

```
### Relationship between variables, categorical by categorical [SKIP]

Tables above are pretty ugly

Use the `spread()` function from `tidyr` package to create table with one variable as columns and the other variable as rows

- The variable you place in `spread()` will be columns
- We learn `spread()` function next week

```{r, results="hide"}
hsls_small %>% group_by(s3classes) %>% count(s3clglvl) %>% 
  spread(s3classes, n)

hsls_small %>% group_by(s3classes) %>% count(s3clglvl) %>% 
  as_factor() %>%  spread(s3classes, n)
hsls_small %>% group_by(s3classes) %>% count(s3clglvl) %>% 
  as_factor() %>%  spread(s3clglvl, n)
```

### Relationship between variables, categorical by continuous

Investigating relationship between multiple variables is a little tougher when at least one of the variables is continuous

__Conditional mean__ (like regression with continuous Y and one categorical X):

- Shows average values of continous variables within groups
- Groups are defined by your categorical variable(s)

__key to syntax__

- `group_by(categorical_var) %>% summarise_at(.vars = vars(continuous_var)`

### Relationship between variables, categorical by continuous

__Task__

- Calculate mean math score, `x2txmtscor`, for each value of parental education, `x2paredu`

```{r, results="hide"}
#first, investigate parental education [print in console]
hsls_small %>% count(x2paredu) 
```

```{r}
# using dplyr to get average math score by parental education level [print in console]
hsls_small %>% group_by(x2paredu) %>%
    summarise_at(.vars = vars(x2txmtscor),
                 .funs = funs(mean, .args = list(na.rm = TRUE))) 
```

### Relationship between variables, categorical by continuous

__Task__

- Calculate mean math score, `x2txmtscor`, for each value of `x2paredu`

For checking data quality, helpful to calculate other stats besides mean
```{r, results="hide"}
hsls_small %>% group_by(x2paredu) %>% #[print in console]
    summarise_at(.vars = vars(x2txmtscor),
                 .funs = funs(mean, min, max, .args = list(na.rm = TRUE)))

```
Always Investigate presence of missing/skip values
```{r, results="hide"}
hsls_small %>% filter(x2paredu<0) %>% count(x2paredu)
hsls_small %>% filter(x2txmtscor<0) %>% count(x2txmtscor)
```

Replace `-8` with `NA` and re-calculate conditional stats
```{r, results="hide"}
hsls_small %>% 
  mutate(x2paredu_na=ifelse(x2paredu<0,NA,x2paredu),
         x2txmtscor_na=ifelse(x2txmtscor<0,NA,x2txmtscor)) %>% 
  group_by(x2paredu_na) %>%
  summarise_at(.vars = vars(x2txmtscor_na),
               .funs = funs(mean, min, max, .args = list(na.rm = TRUE))) %>%
  as_factor()

hsls_small %>% count(s3classes,s3clglvl) %>% as_factor
```
### Student exercise

Can use same approach to calculate conditional mean by multiple `group_by()` variables

- Just add additional variables within `group_by()`

1. Calculate mean math test score (`x2txmtscor`), for each combination of parental education (`x2paredu`) and sex (`x2sex`).  


### Student exercise solution  
1. Calculate mean math test score (`x2txmtscor`), for each combination of parental education (`x2paredu`) and sex (`x2sex`)
```{r, results="hide"}
#hsls_small %>% count(x2sex)

hsls_small %>%
  group_by(x2paredu,x2sex) %>%
  summarise_at(.vars = vars(x2txmtscor),
               .funs = funs(mean, .args = list(na.rm = TRUE))) %>%
  as_factor()
```

## Guidelines for EDA

### Guidelines for "EDA for data quality"

Assme that your goal in "EDA for data quality" is to investigate "input" data sources and create "analysis variables"

- Usually, your analysis dataset will incorporate multiple sources of input data, including data you collect (primary data) and/or data collected by others (secondary data)

While this is not a linear process, these are the broad steps I follow 

1. Understand how input data sources were created
    - e.g., when working with survey data, have survey questionnaire and codebooks on hand
1. For each input data source, identify the "unit of analysis" and which combination of variables uniquely identify observations
1. Investigate patterns in input variables
1. Create analysis variable from input variable(s)
1. Verify that analysis variable is created correctly through descriptive statistics that compare values of input variable(s) against values of the analysis variable

__Always be aware of missing values__

- They will not always be coded as `NA` in input variables

### "Unit of analysis" and which variables uniquely identify observations

"Unit of analysis" refers to "what does each observation represent" in an input data source

- If each obs represents a student, you have "student level data"
- If each obs represents a student-course, you have "student-course level data"
- If each obs represents a school, you have "school-level data"
- If each obs represents a school-year, you have "school-year level data"

How to identify unit of analysis

- data documentation
- investigating the data set

We will go over syntax for identifying unit of analysis in subsequent weeks

### Rules for variable creation

Rules I follow for variable creation

1. \medskip Never modify "input variable"; instead create new variable based on input variable(s)
    - Always keep input variables used to create new variables
1. Investigate input variable(s) and relationship between input variables
1. Developing a plan for creation of analysis variable
    - e.g., for each possible value of input variables, what should value of analysis variable be?
1. Write code to create analysis variable
1. Run descriptive checks to verify new variables are constructed correctly
    - Can "comment out" these checks, but don't delete them
1. Document new variables with notes and labels

### Rules for variable creation

__Task__: 

- Create analysis for variable ses qunitile called `sesq5` based on `x4x2sesq5` that converts negative values to `NAs`

```{r, results="hide"}
#investigate input variable
hsls_small %>% select(x4x2sesq5) %>% var_label()
hsls_small %>% select(x4x2sesq5) %>% val_labels()
hsls_small %>% select(x4x2sesq5) %>% count(x4x2sesq5)
hsls_small %>% select(x4x2sesq5) %>% count(x4x2sesq5) %>% as_factor()

#create analysis variable
hsls_small_temp <- hsls_small %>% 
  mutate(sesq5=ifelse(x4x2sesq5==-8,NA,x4x2sesq5)) # approach 1
hsls_small_temp <- hsls_small %>% 
  mutate(sesq5=ifelse(x4x2sesq5<0,NA,x4x2sesq5)) # approach 2

#verify
hsls_small_temp %>% group_by(x4x2sesq5) %>% count(sesq5)
```


## Skip patterns in survey data

### What are skip patterns

Pretty easy to create an analysis variable based on a single input variable

Harder to create analysis variables based on multiple input variables

- When working with survey data, even seemingly simple analysis variables require multiple input variables due to "skip patterns"

What are "skip patterns"? 

- Response on a particular survey item determines whether respondent answers some set of subsequent questions
- What are some examples of this?

Key to working with skip patterns

- Have the survey questionnaire on hand
- Sometimes it appears that analysis variable requires only one input variable, but really depends on several input variables because of skip patterns
    - Don't just blindly turn "missing" and "skips" from survey data to `NAs` in your analysis variable
    - Rather, trace why these "missing" and "skips" appear and decide how they should be coded in your analysis variable


# Problem Set 7 

### Overview of problem set due next week

__Assignment__: 

- create GPA from postsecondary transcript student-course level data

__Data source__: [National Longitudinal Study of 1972 (NLS72)](https://nces.ed.gov/surveys/nls72/)

- Follows 12th graders from 1972
    - Base year: 1972
    - Follow-up surveys in:  1973, 1974, 1976, 1979, 1986
    - Postsecondary transcripts collected in 1984

__Why use such an old survey for this assignment?__

- NLS72 predates data privacy agreements; transcript data publicly available

__What we do to make assignment more manageable__

- last week's problem set created the input var: numgrade
- we give you some hints/guidelines
- but you are responsible for developing plan to create GPA vars and for executing plan (rather than us giving you step-by-step quations)

__Why this assignment?__

1. Give you more practice investigating data, cleaning data, creating variables that require processing across rows
1. Real world example of "simple" task with complex data management needs




<!-- ## Skip patterns in survey data -->

<!-- ### What are skip patterns -->

<!-- Pretty easy to create an analysis variable based on a single input variable -->

<!-- Harder to create analysis variables based on multiple input variables -->

<!-- - When working with survey data, even seemingly simple analysis variables require multiple input variables due to "skip patterns" -->

<!-- What are "skip patterns"?  -->

<!-- - Response on a particular survey item determines whether respondent answers some set of subsequent questions -->
<!-- - What are some examples of this? -->

<!-- Key to working with skip patterns -->

<!-- - Have the survey questionnaire on hand -->
<!-- - Sometimes it appears that analysis variable requires only one input variable, but really depends on several input variables because of skip patterns -->
<!--     - Don't just blindly turn "missing" and "skips" from survey data to `NAs` in your analysis variable -->
<!--     - Rather, trace why these "missing" and "skips" appear and decide how they should be coded in your analysis variable -->


<!-- ### Creating analysis variables in the presence of skip patterns -->

<!-- Task: Create a measure  of "level" of postsecondary institution attended in 2013 from HSLS:09 survey data -->

<!-- - "level" is highest award-level of the postsecondary institution -->
<!--     - e.g., if highest award is associate's degree (a two-year degree), then `level==2' -->
<!-- - The measure, `pselev2013`, should have following [non-missing] values: -->
<!--     1. Not attending postsecondary education institution -->
<!--     2. Attending a 2-year or less-than-2-year institution -->
<!--     3. Attending 4-year or greater-than-4year institution -->

<!-- Background info: -->

<!-- - In "2013 Update" of HSLS:09, students asked about college attendance -->
<!--     - Variables from student responses to "2013 Update" have prefix `s3` -->
<!-- - Survey questionnaire for 2013 update can be found [HERE](https://nces.ed.gov/surveys/hsls09/pdf/2013_Student_Parent_Questionnaire.pdf) -->
<!-- - The "online codebook" website [HERE](https://nces.ed.gov/onlinecodebook) has info about specific variables -->
<!-- - Measure has 3 input variables [usually must figure this out yourself]: -->
<!--     1. `x3sqstat`: "X3 Student questionnaire status" -->
<!--     1. `s3classes`: "S3 B01A Taking postsecondary classes as of Nov 1 2013" -->
<!--     1. `s3clglvl`: "S3 Enrolled college IPEDS level" -->

<!-- ```{r, results="hide"} -->
<!-- hsls_small %>% select(x3sqstat,s3classes,s3clglvl) %>% var_label() -->
<!-- ``` -->
<!-- <!-- You won't have time to complete this task, but develop a plan for the task and get as far as you can --> -->

<!-- ### Creating analysis variables in the presence of skip patterns -->

<!-- Step 1a: Investigate each input variable separately -->
<!-- ```{r, results="hide"} -->
<!-- #variable labels -->
<!-- hsls_small %>% select(x3sqstat,s3classes,s3clglvl) %>% var_label() -->

<!-- hsls_small %>% count(x3sqstat) -->
<!-- hsls_small %>% count(x3sqstat) %>% as_factor() -->

<!-- hsls_small %>% count(s3classes) -->
<!-- hsls_small %>% count(s3classes) %>% as_factor() -->

<!-- hsls_small %>% count(s3clglvl) -->
<!-- hsls_small %>% count(s3clglvl) %>% as_factor() -->

<!-- ``` -->
<!-- ### Creating analysis variables in the presence of skip patterns -->

<!-- Step 1b: Investigate relationship between input variables -->
<!-- ```{r, results="hide"} -->
<!-- #x3sqstate and s3classes -->
<!-- hsls_small %>% group_by(x3sqstat) %>% count(s3classes)  -->
<!-- hsls_small %>% group_by(x3sqstat) %>% count(s3classes) %>% as_factor() -->

<!-- hsls_small %>% filter(x3sqstat==8) %>% count(s3classes) -->
<!-- hsls_small %>% filter(x3sqstat==8) %>% count(s3classes==-8) -->
<!-- hsls_small %>% filter(x3sqstat !=8) %>% count(s3classes) -->

<!-- #x3sqstate, s3classes and s3clglvl -->
<!-- hsls_small %>% group_by(s3classes) %>% count(s3clglvl)  -->
<!-- hsls_small %>% group_by(s3classes) %>% count(s3clglvl) %>% as_factor() -->

<!-- #add filter for whether student did not respond to X3 questionnaire -->
<!-- hsls_small %>% filter(x3sqstat==8) %>% group_by(s3classes) %>% count(s3clglvl)  -->
<!-- hsls_small %>% filter(x3sqstat !=8) %>% group_by(s3classes) %>% count(s3clglvl) -->

<!-- #continued on the next page -->
<!-- ``` -->


<!-- ### Creating analysis variables in the presence of skip patterns -->

<!-- Step 1b: Investigate relationship between input variables continued... -->
<!-- ```{r} -->
<!-- #add filter for s3classes is "missing" [-9] -->
<!-- hsls_small %>% filter(x3sqstat !=8,s3classes==-9) %>% group_by(s3classes) %>% -->
<!--   count(s3clglvl) -->
<!-- hsls_small %>% filter(x3sqstat !=8,s3classes!=-9) %>% group_by(s3classes) %>% -->
<!--   count(s3clglvl) -->

<!-- #add filter for s3classes equal to "no" or "don't know" -->
<!-- hsls_small %>% filter(x3sqstat !=8,s3classes!=-9, s3classes %in% c(2,3)) %>% -->
<!--   group_by(s3classes) %>% count(s3clglvl) -->
<!-- hsls_small %>% filter(x3sqstat !=8,s3classes!=-9, s3classes %in% c(2,3)) %>%  -->
<!--   group_by(s3classes) %>% count(s3clglvl) %>% as_factor() -->

<!-- hsls_small %>% filter(x3sqstat !=8,s3classes!=-9, s3classes==1) %>%  -->
<!--   group_by(s3classes) %>% count(s3clglvl) -->
<!-- hsls_small %>% filter(x3sqstat !=8,s3classes!=-9, s3classes==1) %>%  -->
<!--   group_by(s3classes) %>% count(s3clglvl) %>% as_factor() -->
<!-- ``` -->


