---
title: "Problem Set 1"
author: "insert your name here"
date: "insert date here"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to Data Management and Manipulation Using R!

In this problem set, you will load data from url, or absolute/relative filepath.

    - Note: Change author to your name & date to current date 


### Step 1:

- Go to your rclass folder and open your saved lecture1.2.Rmd file (hint: in lecture1 folder)  
- Open the lecture1.2.Rmd file in RStudio  
- At the top of the lecture1.2.Rmd file, before the "# R basics", type your first and last name (e.g. "Patricia Martin")  
- Now select the "Knit" tab (icon with blue yarn ball) or scroll down and "Knit to PDF"  
- The goal of this assignment is to make sure you are able to knit to a PDF without running into errors.  


### Step 2:  

**Question 1**: Load .Rdata directly with url  

- Make sure to view current/working directory first `getwd()` 
- Url to data https://github.com/ozanj/rclass/raw/master/data/recruiting/recruit_ps1_small.Rdata  
- Hint: to load .Rdata use `load()` function and "url" because you are using a link 
- E.g. `load(url("link"))`  

```{r}

```


**Question 2**: Set your working directory and use absolute filepath to load data  

- Download .Rdata using link from above  
- Save .Rdata file in the recruiting folder (rclass/data/recruiting)  
- View your current/working directory `getwd()` 
- Use `setwd()` function to set your working directory to where .Rdata file is saved (rclass/data/recruiting)  
- When setting working directory, use the absolute filepath  
- Load the .Rdata file  `load(name.Rdata)`
- Using the `names(data frame name)` function, report the names in this dataset  (Hint: data frame name is df_recruiting)  

```{r}

```

**Question 3**: Set your working directory and use relative filepath to load data  

- View your current/working directory `getwd()` 
- Use `setwd()` function to set your working directory to where .Rdata file is saved (rclass/data/recruiting)  
- When setting working directory, use the relative filepath  
- Load the .Rdata file  `load()`
- Report the names in this dataset (Hint: data frome name is df_recruiting)  

```{r}

```

**Question 4**: Load dataset using relative filepath without first changing working directory  

- View your current/working directory `getwd()`  
- Load data without setting directory `load(filepath/data.Rdata)`

```{r}

```

**Knit to pdf** select the "Knit" tab (icon with blue yarn ball) or scroll down and "Knit to PDF"  

- Submit both .Rmd and pdf   
- Use this naming convention "lastname_firstname_ps#"
