---
title: "Lecture 3: Variable creation, variable attributes, and pipes"
subtitle:  "EDUC 263: Managing and Manipulating Data Using R"
author: Ozan Jaquette
date: 
fontsize: 8pt
classoption: dvipsnames  # for colors
urlcolor: blue
output:
  beamer_presentation:
    keep_tex: true
    toc: true
    slide_level: 3
    theme: default # AnnArbor # push to header?
    #colortheme: "dolphin" # push to header?
    #fonttheme: "structurebold"
    highlight: default # Supported styles include "default", "tango", "pygments", "kate", "monochrome", "espresso", "zenburn", and "haddock" (specify null to prevent syntax highlighting); push to header
    df_print: tibble #default # tibble # push to header?    
    latex_engine: xelatex #  Available engines are pdflatex [default], xelatex, and lualatex; The main reasons you may want to use xelatex or lualatex are: (1) They support Unicode better; (2) It is easier to make use of system fonts.
    includes:
      in_header: ../beamer_header.tex
      #after_body: table-of-contents.txt 
---

<!-- CAN THIS BE MOVED TO SOME KIND OF HEADER FILE? --> 

```{r, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
#knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```

# Introduction/logistics

### Libraries we will use today

```{r}
library(tidyverse)
```

### Data we will use today

Data on off-campus recruiting events by public universities

```{r}
rm(list = ls()) # remove all objects

#load dataset with one obs per recruiting event
load("../../data/recruiting/recruit_event_somevars.Rdata")

#load dataset with one obs per high school
load("../../data/recruiting/recruit_school_somevars.Rdata")

load("../../data/prospect_list/western_washington_college_board_list.RData")

```
Object \hlgc{df\_event\}

- One observation per university, recruiting event

Object \hlgc{df\_event\}

- One observation per high school (visited and non-visited)

# Factors

## Review data types and structures

### Review data types

Primary __data types__ in R:

- numeric (integer & double)
- character
- logical

R CODE CHUNK WITH EXAMPLES

### Review data structures: vectors

Primary __data structures__ in R are __vectors__ and __lists__

\medskip A __vector__ is a collection of values

- each value in a vector is an __element__
- all elements within vector must have same __data type__

```{r}
a <- c(1,2,3)
a
str(a)
```

You can assign __names__ to elements of a vector, thereby creating a __named vector__

```{r}
b <- c(v1=1,v2=2,v3=3)
b
str(b)
```

### Review data structures: lists

Like vectors, __lists__ are objects that contain __elements__; However, __data type__ can differ across elements within a list; an element of a list can be another list


Examples of lists:
```{r}
list_a <- list(1,2,"apple")
str(list_a)
list_b <- list(1, c("apple", "orange"), list(1, 2, 3))
str(list_b)
```
### Review data structures: lists

Like vectors, elements within a list can be named, thereby creating a __named list__

```{r}
str(list_b) # not named

list_c <- list(v1=1, v2=c("apple", "orange"), v3=list(1, 2, 3))
str(list_c) # named
```

### Review data structures: a data frame is a list

A __data frame__ is a list with the following characteristics:

- All the elements must be __vectors__ with the same __length__
- Data frames are __augmented lists__ because they have additional __attributes__ [described later]

```{r}
list_d <- list(col_a = c(1,2,3), col_b = c(4,5,6), col_c = c(7,8,9))
typeof(list_d)
str(list_d)

df_a <- data.frame(col_a = c(1,2,3), col_b = c(4,5,6), col_c = c(7,8,9))
typeof(df_a)
str(df_a)
```
## Attributes and augmented vectors

### Atomic vectors versus augmented vectors

__Atomic vectors__ [our focus so far]
- [(See figure)](http://r4ds.had.co.nz/diagrams/data-structures-overview.png)
- I think of atomic vectors as "just the data"
- Atomic vectors are the building blocks for augmented vectors

Augmented vectors

- __Augmented vectors__ are atomic vectors with additional __atributes__ attached

__Attributes__

- __Attributes__ are additional "metadata" that can be attached to any object (e.g., vector or list)
- Important attributes in R:
    - __Names__: name the elements of a vector (e.g., variable names)
    - __Class__: How object should be treated by object oriented programming language [discussed below]

Main takaway:

- Augmented vectors are atomic vectors (just the data) with additional attributes attached

### Attributes in vectors

```{r}
vector1 <- c(1,2,3,4)
vector1
attributes(vector1)

vector2 <- c(a = 1, b= 2, c= 3, d = 4)
vector2
attributes(vector2)
```


### Attributes in lists

```{r}
list1 <- list(c(1,2,3), c(4,5,6))
str(list1)
attributes(list1)

list2 <- list(col_a = c(1,2,3), col_b = c(4,5,6))
str(list2)
attributes(list2)

list3 <- data.frame(col_a = c(1,2,3), col_b = c(4,5,6))
str(list3)
attributes(list3)

```
### Object class

```{r}
vector1 <- c(1,2,3,4)
vector1
typeof(vector1)
class(vector1)
attributes(vector1)

vector2 <- c(a = 1, b= 2, c= 3, d = 4)
vector2
attributes(vector2)
typeof(vector2)
class(vector2)


```

# FACTORS

### Factors

__Factors__ are used to display categorical data (e.g., marital status)

- A factor is an __augmented vector__ built by attaching a "levels" attribute to an (atomic) integer vectors

The \hlgc{str()} function is useful for identifying which variables are factors. Let's examine the factor variable \hlgc{ethn\_code}

```{r}
typeof(wwlist$ethn_code)
class(wwlist$ethn_code)
str(wwlist$ethn_code)
```
Note that \hlgc{ethn\_code} has \hlgc{type=integer} and \hlgc{class=factor} because the variable has a "levels" attribute

```{r, results='hide'}
attributes(wwlist$ethn_code)
```

Main takeaway:

- The underlying data are integers but the levels attribute is used to display the data.



### Working with factor variables

```{r, results='hide'}
attributes(wwlist$ethn_code)
```
Refer to categories of a factor by the values of the level attribute rather than the underlying values of the variable

```{r}
count(filter(wwlist,ethn_code==11))

count(filter(wwlist,ethn_code=="White"))
```

If you want to refer to underlying values, then apply \hlgc{as.integer()} function to the factor variable
```{r}
count(filter(wwlist,as.integer(ethn_code)==11))
```
### How to identify the variable values associated with factor levels

MAYBE CUT THIS SLIDE IF YOU CAN'T DO THIS WITHOUT PIPES
```{r}

wwlist %>% count(psat_range) %>% as_factor()
```
```{r}
count(filter(wwlist,as.integer(psat_range)==4))
count(filter(wwlist,psat_range=="1270-1520"))
```


### Some in-class exercise involving factors

```{r}
str(wwlist)
```
### Creating factors [from integer vectors]

Factors are just integer vectors with level attributes attached to them. So, to create a factor:

1. create a vector for the underlying data
1. create a vector that has level attributes
3. Attach levels to the data using the \hlgc{factor()} function

```{r}
a1 <- c(1,1,1,0,1,1,0) #a vector of data
a2 <- c("zero","one") #a vector of labels

#attach labels to values
a3 <- factor(a1, labels = a2)
a3
str(a3)

```

Note: By default, \hlgc{factor()} function attached "zero" to the lowest value of vector \hlgc{a1} because "zero" was the first element of vector \hlgc{a2}

### Creating factors [from integer vectors]

Let's turn an integer variable into a factor variable in the \hlgc{wwlist} data frame

Create integer version of \hlgc{sex}
```{r}
wwlist$sex_int <- as.integer(wwlist$sex)
str(wwlist$sex_int)
#wwlist %>% count(sex) %>% as_factor()
```
Assume we know that 1=female, 2=male, 3=unknown

Assign levels to values of integer variable
```{r}
wwlist$sex_int <- factor(wwlist$sex_int, labels=c("female","male","unknown"))
str(wwlist$sex_int)
str(wwlist$sex)
```

### Create factors [from string variables]

To create a factor variable from string variable

1. create a character vector containing underlying data
1. create a vector containing valid levels
3. Attach levels to the data using the \hlgc{factor()} function

```{r}
#underlying data: months my fam is born
x1 <- c("Jan", "Aug", "Apr", "Mar")
#create vector with valid levels
month_levels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
#attach levels to data
x2 <- factor(x1, levels = month_levels)
```
Note how attributes differ
```{r}
str(x1)
str(x2)
```
Sorting differs
```{r}
sort(x1)
sort(x2)
```

### Create factors [from string variables]

Let's create a character version of variable \hlgc{sex} and then turn it into a factor

```{r}
#Create character version of sex
wwlist$sex_char <- as.character(wwlist$sex)

#investigate character variable
str(wwlist$sex_char)
table(wwlist$sex_char)

#create new variable that assigns levels
sex_fac <- factor(wwlist$sex_char, levels = c("F","M","U"))
str(wwlist$sex_char)
```
How the \hlgc{levels} argument works when underlying data is character

- Matches value of underlying data to value of the level attribute
- Converts underlying data to integer, with level attribute attached

\medskip See chapter 15 of Wickham for more on factors (e.g., modifying factor order, modifying factor levels)

### Substantial exercise on using/creating factors, using either df_school or df_event datasets


